---
title: "IST772– Problem Set 1"
author: "Sathish Kumar Rajendiran"
output: pdf_document
---

1. I did this homework by myself, with help from the book and the professor. 

# Chapter 1, Exercise 1

_Using the material from this chapter and possibly other information that you look up, write a brief definition of these terms in your own words: mean (aka average), median, mode, variance, standard deviation, histogram, normal distribution, and Poisson distribution. (1 point for each definition)_

```{r}
cat("\n ****************************************************************************")
```

* Mean: 
          To find the mean or average of all the values in a data set, simply add up all the numbers and divide by total number of values from the data set. 

* Median: 
          The median or the halfway value of a data set is the middle data value when the data are put into ascending order. 
          If the sample size n is odd, then the median is the middle value. 
          If the sample size n is even, then the median is the mean of the middle data values. 
* Mode:
          The mode or the modal value of the data is the data value that occurs the greatest frequency.

* Variance:
          variance is approximately the mean of the squared diviations or simply standard deviation squared.

* Standard deviation:
           Standard deviation can be interpreted as the typical differnce between a data value and the sample mean for a given data set or simply the square root of the variance.

* Histogram:
          Graphical representation of the numerical data distribution (spread) using bars with varying heights. Each bar groups number into ranges. Taller bars show that more data falls in that range.
          
* Normal distribution:
          Distribution of a continuous data with symmetric tails and gradual curves toward a peak in the middle is called normal distribution or simply "the bell curve".

* Poisson distribution:
         When the distribution of discrete data represented by different shapes at differnt levels of lamdba ("mean" value) is called Poisson distribution. In this all the observations are positive.

```{r}
cat("\n ****************************************************************************")
```

# Chapter 1, Exercise 3

_Use the data() function to get a list of the data sets that are included with the basic installation of R: just type “data()” at the command line and press enter._ 

```{r }
data()
```

Data sets in package ‘datasets’:

AirPassengers                                   Monthly Airline Passenger Numbers 1949-1960
BJsales                                         Sales Data with Leading Indicator
BJsales.lead (BJsales)                          Sales Data with Leading Indicator
BOD                                             Biochemical Oxygen Demand
CO2                                             Carbon Dioxide Uptake in Grass Plants
ChickWeight                                     Weight versus age of chicks on different diets
DNase                                           Elisa assay of DNase
EuStockMarkets                                  Daily Closing Prices of Major European Stock Indices, 1991-1998
Formaldehyde                                    Determination of Formaldehyde
HairEyeColor                                    Hair and Eye Color of Statistics Students
Harman23.cor                                    Harman Example 2.3
Harman74.cor                                    Harman Example 7.4
Indometh                                        Pharmacokinetics of Indomethacin
InsectSprays                                    Effectiveness of Insect Sprays
JohnsonJohnson                                  Quarterly Earnings per Johnson & Johnson Share
LakeHuron                                       Level of Lake Huron 1875-1972
LifeCycleSavings                                Intercountry Life-Cycle Savings Data
Loblolly                                        Growth of Loblolly pine trees
Nile                                            Flow of the River Nile
Orange                                          Growth of Orange Trees
OrchardSprays                                   Potency of Orchard Sprays
PlantGrowth                                     Results from an Experiment on Plant Growth
Puromycin                                       Reaction Velocity of an Enzymatic Reaction
Seatbelts                                       Road Casualties in Great Britain 1969-84
Theoph                                          Pharmacokinetics of Theophylline
Titanic                                         Survival of passengers on the Titanic
ToothGrowth                                     The Effect of Vitamin C on Tooth Growth in Guinea Pigs
UCBAdmissions                                   Student Admissions at UC Berkeley
UKDriverDeaths                                  Road Casualties in Great Britain 1969-84
UKgas                                           UK Quarterly Gas Consumption
USAccDeaths                                     Accidental Deaths in the US 1973-1978
USArrests                                       Violent Crime Rates by US State
USJudgeRatings                                  Lawyers' Ratings of State Judges in the US Superior Court
USPersonalExpenditure                           Personal Expenditure Data
UScitiesD                                       Distances Between European Cities and Between US Cities
VADeaths                                        Death Rates in Virginia (1940)
WWWusage                                        Internet Usage per Minute
WorldPhones                                     The World's Telephones
ability.cov                                     Ability and Intelligence Tests
airmiles                                        Passenger Miles on Commercial US Airlines, 1937-1960
airquality                                      New York Air Quality Measurements
anscombe                                        Anscombe's Quartet of 'Identical' Simple Linear Regressions
attenu                                          The Joyner-Boore Attenuation Data
attitude                                        The Chatterjee-Price Attitude Data
austres                                         Quarterly Time Series of the Number of Australian Residents
beaver1 (beavers)                               Body Temperature Series of Two Beavers
beaver2 (beavers)                               Body Temperature Series of Two Beavers
cars                                            Speed and Stopping Distances of Cars
chickwts                                        Chicken Weights by Feed Type
co2                                             Mauna Loa Atmospheric CO2 Concentration
crimtab                                         Student's 3000 Criminals Data
discoveries                                     Yearly Numbers of Important Discoveries
esoph                                           Smoking, Alcohol and (O)esophageal Cancer
euro                                            Conversion Rates of Euro Currencies
euro.cross (euro)                               Conversion Rates of Euro Currencies
eurodist                                        Distances Between European Cities and Between US Cities
faithful                                        Old Faithful Geyser Data
fdeaths (UKLungDeaths)                          Monthly Deaths from Lung Diseases in the UK
freeny                                          Freeny's Revenue Data
freeny.x (freeny)                               Freeny's Revenue Data
freeny.y (freeny)                               Freeny's Revenue Data
infert                                          Infertility after Spontaneous and Induced Abortion
iris                                            Edgar Anderson's Iris Data
iris3                                           Edgar Anderson's Iris Data
islands                                         Areas of the World's Major Landmasses
ldeaths (UKLungDeaths)                          Monthly Deaths from Lung Diseases in the UK
lh                                              Luteinizing Hormone in Blood Samples
longley                                         Longley's Economic Regression Data
lynx                                            Annual Canadian Lynx trappings 1821-1934
mdeaths (UKLungDeaths)                          Monthly Deaths from Lung Diseases in the UK
morley                                          Michelson Speed of Light Data
mtcars                                          Motor Trend Car Road Tests
nhtemp                                          Average Yearly Temperatures in New Haven
nottem                                          Average Monthly Temperatures at Nottingham, 1920-1939
npk                                             Classical N, P, K Factorial Experiment
occupationalStatus                              Occupational Status of Fathers and their Sons
precip                                          Annual Precipitation in US Cities
presidents                                      Quarterly Approval Ratings of US Presidents
pressure                                        Vapor Pressure of Mercury as a Function of Temperature
quakes                                          Locations of Earthquakes off Fiji
randu                                           Random Numbers from Congruential Generator RANDU
rivers                                          Lengths of Major North American Rivers
rock                                            Measurements on Petroleum Rock Samples
sleep                                           Student's Sleep Data
stack.loss (stackloss)                          Brownlee's Stack Loss Plant Data
stack.x (stackloss)                             Brownlee's Stack Loss Plant Data
stackloss                                       Brownlee's Stack Loss Plant Data
state.abb (state)                               US State Facts and Figures
state.area (state)                              US State Facts and Figures
state.center (state)                            US State Facts and Figures
state.division (state)                          US State Facts and Figures
state.name (state)                              US State Facts and Figures
state.region (state)                            US State Facts and Figures
state.x77 (state)                               US State Facts and Figures
sunspot.month                                   Monthly Sunspot Data, from 1749 to "Present"
sunspot.year                                    Yearly Sunspot Data, 1700-1988
sunspots                                        Monthly Sunspot Numbers, 1749-1983
swiss                                           Swiss Fertility and Socioeconomic Indicators (1888) Data
treering                                        Yearly Treering Data, -6000-1979
trees                                           Diameter, Height and Volume for Black Cherry Trees
uspop                                           Populations Recorded by the US Census
volcano                                         Topographic Information on Auckland's Maunga Whau Volcano
warpbreaks                                      The Number of Breaks in Yarn during Weaving
women                                           Average Heights and Weights for American Women

Data sets in package ‘ggplot2’:

diamonds                                        Prices of over 50,000 round cut diamonds
economics                                       US economic time series
economics_long                                  US economic time series
faithfuld                                       2d density estimate of Old Faithful data
luv_colours                                     'colors()' in Luv space
midwest                                         Midwest demographics
mpg                                             Fuel economy data from 1999 to 2008 for 38 popular models of cars
msleep                                          An updated and expanded version of the mammals sleep dataset
presidential                                    Terms of 11 presidents from Eisenhower to Obama
seals                                           Vector field of seal movements
txhousing                                       Housing sales in TX

_Choose a data set from the list that contains at least one numeric variable--for example, the Biochemical Oxygen Demand (BOD) data set._
_Use the summary() command to summarize the variables in the data set you selected--for example, summary(BOD). (1 pt)_
_Write a brief description of the mean and median of each numeric variable in the data set. (1 pt for each value)_
_Make sure you define what a “mean” and a “median” are, that is, the technical definition and practical meaning of each of these quantities. (1 pt for each definition)_

```{r}
cat("\n ****************************************************************************")
#Biochemical Oxygen Demand
# ?BOD   #view definition of the BOD dataset; This data set has two numeric variables Time and demand and 6 observations.
myBOD <- BOD

cat("\n Structure of the Biochemical Oxygen Demand data set is:\n")
str(myBOD) # analyze the structure of the data set. 
# View(myBOD)  # view the data set

cat("\n Summary of the Biochemical Oxygen Demand data set is:\n")
summary(myBOD)  # Summarize the variables from the dataset 
  # Time            demand  
  # Mean   :3.667   Mean   :14.83  

#calculate Mean and Median of "Time" of the variable
mean_time <- mean(myBOD$Time)
median_time <- median(myBOD$Time)
cat("\n mean of the Time variable is:",mean_time)
cat("\n median of the Time variable is:",median_time)

#calculate Mean and Median of "demand" of the variable
mean_demand <- mean(myBOD$demand)
median_demand <- median(myBOD$demand)

cat("\n mean of the demand variable is:",mean_demand)
cat("\n median of the demand variable is:",median_demand,"\n")
cat("\n ****************************************************************************")
cat("\n Mean:\n")
cat("    Mean is the average value from the observations.It is calculated by sum of all  
    observation values and divide by the total number of observations.
 Practical Meaning:
    In this dataset there are two variables (Time and demand) having 6 observations of each.")
cat("\n Time:")
time <- myBOD$Time
S_Time <- sum(time)
n_Time <- length(time)
avg_time <- S_Time/n_Time

cat("\n    Sum of the Time variable is:",S_Time)
cat("\n    Count of the Time variable is:",n_Time)
cat("\n    Mean of the Time variable is:",avg_time)

cat("\n Demand:")
demand <- myBOD$demand
S_demand <- sum(demand)
n_demand <- length(demand)
avg_demand <- S_demand/n_demand
cat("\n    Sum of the Demand variable is:",S_demand)
cat("\n    Count of the Demand variable is:",n_demand)
cat("\n    Mean of the Demand variable is:",avg_demand)
cat("\n ****************************************************************************")
cat("\n Median:\n")
cat("    The median or the halfway value of a data set is the middle data value when the data are put into ascending order. 
    If the sample size n is odd, then the median is the middle value. 
    If the sample size n is even, then the median is the mean of the middle data values.
 Practical Meaning:
    In this dataset there are two variables (Time and demand) having 6 (even number of sample size) observations of each.So,the median would be calculated by 
    taking the mean of middle data values. In this case 3 and 4 position values")
cat("\n Time:")
sorted_time <- sort(time,decreasing = FALSE)
sum_middle_values_time <- sum(sorted_time[3:4])
n1 <- length(sorted_time[3:4])
cat("\n    sorted observations of time variable values:",sorted_time)
cat("\n    sum of middle 2 values of sorted time variable:",sum_middle_values_time)
cat("\n    median of time variable:",sum_middle_values_time/n1)

cat("\n Demand:")
sorted_demand <- sort(demand,decreasing = FALSE)
sum_middle_values_demand <- sum(sorted_demand[3:4])
n2 <- length(sorted_demand[3:4])
cat("\n    sorted observations of demand variable values:",sorted_demand)
cat("\n    sum of middle 2 values of sorted demand variable:",sum_middle_values_demand)
cat("\n    median of demand variable:",sum_middle_values_demand/n2)
cat("\n ****************************************************************************")
```


# Chapter 1, Exercise 4

_As in the previous exercise, use the data() function to get a list of the data sets that are included with the basic installation of R. Choose a data set and pick out one variable, for example, the LakeHuron data set (levels of Lake Huron in the years 1875 through 1972)._
_Use the hist() command to create a histogram of the variable--for example, hist(LakeHuron). (2 pts)_
_Describe the shape of the histogram in words. (2 pts)_
_Which of the distribution types do you think these data fit most closely (e.g., normal, Poisson). (2 pts)_
_Speculate on why your selected data may fit that distribution. (2 pts)_ 

```{r}
library("ggplot2")
library("hrbrthemes")
cat("\n ****************************************************************************\n")
# Level of Lake Huron 1875–1972
# ?LakeHuron # Annual measurements of the level, in feet, of Lake Huron 1875–1972. A time series of length 98.
myLakeHuron <- data.frame(LakeHuron)
str(myLakeHuron)
summary(myLakeHuron)
# View(myLakeHuron)
# sd(myLakeHuron$LakeHuron)
# mean(myLakeHuron$LakeHuron)

#calculate Mean and Median of "Time" of the variable
mean_Huron <- mean(myLakeHuron$LakeHuron)
median_Huron <- median(myLakeHuron$LakeHuron)
sd_Huron <- sd(myLakeHuron$LakeHuron)
v_Huron <- var(myLakeHuron$LakeHuron)
cat("\n mean of the LakeHuron variable is:",mean_Huron)
cat("\n median of the LakeHuron variable is:",median_Huron)
cat("\n Standard Deviation of the LakeHuron variable is:",sd_Huron)
cat("\n Variance of the LakeHuron variable is:",v_Huron)
cat("\n ****************************************************************************\n")
```


```{r}
#createa a histogram
hist(myLakeHuron$LakeHuron)
```

```{r}
# create a histogram
# hist(myLakeHuron$LakeHuron)
hist(myLakeHuron$LakeHuron,freq=FALSE,col="#69b3a2",lty="blank",include.lowest = TRUE,main = paste("Histogram of LakeHuron"),xlab="LakeHuron")
lines(density(myLakeHuron$LakeHuron),col="red", lwd=2)
```
 
 
 * Shape of the Histogram:
    This histogram represents a bell shaped curve with one peak value and two low tails on each side.
 
 * Ditribution Type:
    This dataset has continuous data with decimal values; having almost 98 observations, the values are ditributed normally across with symmetric tails and gradual curve towards the peak.Hence, it fits normal distribution pattern closely.
 
 * Conclusion:
    This histogram fits normal distribution closely;becuase, with 98 observations (continuous data), mean and median are almost same 579.0 and 579.1 with standard deviation as 1.and variance 1.7. 
    In addition, the density curve - represents a bell shaped curve almost fitting the values under the curve.So, both the measures of central tendency and dispersion suggests a normal ditribution.

```{r}
cat("\n ****************************************************************************")
```

