---
title: "Week 3 – Sampling Exercise"
author: "Sathish Rajendiran"
---

Instructions: In the asynchronous material for this week, we created and made sense of sampling distributions by creating a simulated population from random data and then drawing samples from that population. By examining large collections of samples, we can understand what to expect when we make inferences from a single sample of real data. In this first live exercise, we’ve used a few lines of code to set up another simulated population: scores on a test of achievement. Achievement tests like this one are calibrated so that the population mean is 100 and the population standard deviation is 10. The diagrams below depict a simulated population of N=100,000 test takers, each of whom scored somewhere between 60 and 140 on the test. Using the diagrams, answer the following questions about the simulated population.

```{r}
set.seed(1234)                                        # Control randomization
testPop <- rnorm(100000, mean=100, sd=10)             # Create simulated population 
testPopMean <- mean(testPop)                          # Calculate mean
sumSq <- sum((testPop-testPopMean)^2)                 # Sum of squares
testPopVar <- sumSq/length(testPop)                   # Variance
testPopSD <- sqrt(testPopVar)                         # Standard deviation
testPopVar
testPopSD
```

1. Here’s a histogram of the population. What is the shape of this distribution and the formal statistical name for that shape:
# a normal (bell-shaped) distribution curve

```{r}
hist(testPop)
```

2. In this next histogram, I’ve marked the 1st, 2nd, and 3rd quartiles. Define the term quartile and estimate the values associated with these quartiles. 
The second quartile (also known as the 50th percentile) has a special name. What is it?

# Quartile indicates that the values are split into 4 equal parts; One quarter on the low end, one quarter on the high end and 2 quarters in the middle.
# 1st Quartile is 93.29; 2nd Quartile is 100.04; 3rd Quartile is 106.74; 4th Quartile is 139.06
# 2nd Quartile (50th percentile is called the Median of the sampling distribution); Value is 100.0449

```{r}
# Markings of quartiles
hist(testPop)                                           
abline(v=quantile(testPop, probs=0.25),col="red")    # First quartile
abline(v=quantile(testPop, probs=0.5),col="green")     # Second quartile
abline(v=quantile(testPop, probs=0.75),col="blue")    # Third quartile

boxplot(testPop,horizontal=TRUE) # boxplot showing quartiles in 4 quadrants
summary(testPop)
mean(testPop)
median(testPop)

```



3. The 1st and 3rd quartiles in that histogram are very close to the middle. Can that be right? What percentage of cases fall below the 1st quartile? Above the 3rd quartile? If you want, you can write a few lines of code to confim the answer. 
# Right. 25% below 1st Quartile and 25% above 3rd Quartile.

```{r}
quantile(testPop, probs=0.25)
quantile(testPop, probs=0.75)
summary(testPop)
sd(testPop)
```

4. By definition, this achievement test has a population standard deviation of 10. I’ve used abline() to mark points that are one and two standard deviations below the mean. 
Examine the histogram below and report what value is two standard deviations below the mean and what value is two standard deviations above the mean:

# mean of the population is 100.02; with standard deviation as ~10,
# the two standard deviation below the mean would be ~80 
# and two standard deviations above the mean would be ~120.


```{r}
# Markings of one and two standard deviations 
hist(testPop)
abline(v=90, col="red")
abline(v=80, col="blue")
abline(v=110, col="red") 
abline(v=120, col="blue")
mean(testPop)

sd(testPop)
```

5. Now I’ve marked some different quantiles, specifically the 0.025 quantile and the 0.975 quantile. 
Estimate where these quantiles fall on the X-axis. What percentage of cases occur between the 0.025 quantile and the 0.975 quantile?

# 0.025 is on the left hand low tail from the central region ; 0.975 is on the right hand low tail from the central region.
# between 0.025 and 0.975 is called the central region that constitues 95% of the data under the curve.


```{r}
# Markings of the central region and tails 
hist(testPop)
abline(v=quantile(testPop, probs=0.025),col="green") 
abline(v=quantile(testPop, probs=0.975),col="green")
quantile(testPop, probs=0.975)
quantile(testPop, probs=0.025)
```

6. The area between the green lines is known as the “central region” and the two areas outside of the green lines are the “tails.”
What percentage of cases fall in the lower tail, that is, below the 0.025 quantile in the histogram above? 
# 2.5%
What percentage of cases fall in the upper tail, that is, above the 0.975 quantile in the histogram above? 
# 2.5%
